On the eletrodynamic of moving bodies

### 2019-06-21 Article_AS.Rmd
M&M

Improvement programs by genetic selection are in place for the most important commercial species [@DRF2000; @Mullin2011] including white spruce (Picea glauca (Moench) Voss) which has a significant economic value because of its geographical distribution and abundance [@Eyre1980; @Abrahamson2015]. Spruce wood is valuable as lumber, material for carpentry products and for pulping because of the characteristics of its fibres [@Zhang2008]. In addition, the plasticity of spruce, enable this species to adapt to various types of soil and climatic conditions [@Nienstaedt1990]. All these characteristics make white spruce a species of great interest in North America and consequently it is heavily reforested in various silvicultural systems in Canada with more than 150 million of natural an hybrid seedlings planted yearly across almost all forested regions which represents a wide range of soils types and a climatic conditions [@Canadian2009]. \

@@

## Sampling
The dataset used in this study came from genetic selection for a series of tests according to provenances [@Beaulieu1996]. The improvement program for this region was led by Quebec’s Ministry of Forests, Wildlife and Parcs, Canada. Experimental plantings were initiated from white spruce seedlings grown in nurseries and planted at two years of age at a regular spacing between trees of 2 m in all directions. This series of tests was established in 1999 by the Canadian Forest Service and is one of the three essential parts of the white spruce breeding program in Quebec. Two sites were selected for sampling: one near Porneuf (Saint-Casimir), Quebec (46.6994, -72.1119), in the maple–basswood ecological zone (warm weather) , and the other one near Cabano (Asselin), Quebec (47.8808, -68.4436), in the balsam fir-yellow birch ecological zone (cool weather) (fig 1). Between 9 and 14 trees per site were sampled from each of the 93 full-sib families, for a total sample of 2365 trees. Tree height and diameter at breast height (DBH, 1.3 m above ground) were recorded in 2015 at the age of 15 years, and a wood increment core was extracted from the south facing side of each tree. Cores were stored in a freezer, conditioned to 7% moisture and cut to 1.68 mm thickness prior to X-ray densitometry analyses (Quintek Measurement Systems, TN) at a resolution of 0.0254 mm, and the data were reported as relative wood density on a dry weight basis. Wood density was calculated as a ring area weighted mean from recorded pith to bark wood density profiles.

@@

To verify the accuracy of dates and determine if there were missing rings, we used COFECHA, which is a dendrochronology software program used for cross-dating and quality control of annual ring measurements [@HOLMES1983]. Thus, as the annual ring separation was carried out by an algorithm, we eliminated the samples (127 samples) whose dating was not correct and could not be modified. Any incomplete or false rings were excluded from the dataset.

## Model Development
A model was developed to describe the patterns of variation in mean annual ring density in the stem in white spruce. Considering the hierarchical structure of the data, with observations on annual rings nested in the cores, cores nested in the trees, and trees nested in blocks, we opted for a mixed-effects modelling approach using nlme package [@NLME2018] in the R statistical programming environment [@R2018]. This type of model also considers the autocorrelation between the observations [@Lindstrom1990]. Thus, the estimates of the parameters can vary randomly around the average values of each level of the data structure (site, block, tree, core) [@Auty2013]. As in the study of @Xiang2014, pith-to-bark wood density profiles in our data were different among cores. A nonlinear equation was fit for each core to describe ring density profiles, producing different parameter estimates for each core. The radial pattern of wood density in white spruce can be described as follows:

@@

where RD is the mean ring density (kg.m^-3^), CA is cambial age (years), $b_0$, $b_1$ and $b_2$ are empirically determined parameters, and $e$ is the residual error term. Parameter $b_0$ denotes the intercept when CA approaches a theoretical value of 0. We noticed that it was impossible to optimize the equation using the same estimated parameters for all tree cores, because the pattern of variation of annual rings average density of each tree is more or less different from the others. We, therefore, used the optim function of the stats package in the R [@R2018] statistical programming environment to find the best parameters iteratively for each tree core . The fit of the model was evaluated by visual analysis of the model residual curves against adjusted values and explanatory variables, fit indices (R^2^) [@Pape1999]. The graphic description in figure 2 of the three parameters of our model makes it possible to understand how each parameter influences our model (Fig 2).

@@

## Determination of the wood density variability index
The variability index is an index to determine the uniformity in the density of pith-to-bark in order to identify tree families that are least sensitive to variations in climatic conditions. The residues of our model were used to obtain the wood density variability index, which was determined using an equation. We subtracted the density residues values of year $n-1$ from year $n$. We were interested more specifically in the difference between the average density residues of annual rings of two consecutive years, no matter whether the difference was positive or negative. The absolute value was used for each difference to get rid of the sign. The index was estimated from the following formula:

@@

## Spearman’s rank correlation coefficient (Spearman’s rho) 
To describe the link between two variables by means of a monotonic function, we used Spearman’s rank correlation coefficient. We calculated this coefficient for the variability index and the different parameters of the model. The elements for which we calculated Spearman’s rho are: the variability index for each of the tree families between two study sites and the three parameters to estimate from the model. The $cor.test$ function in the R [@R2018] statistical programming environment was used to calculate Spearman’s rho. We used a 95 percent confidence interval.

@@

When considering all the cores of trees from the pith-to-bark, ring density was high near the pith and decreased rapidly in the first 5 growth rings. It goes from 430 kg/m^3^ to 370 kg/m^3^ for the Saint-Casimir site and from 455kg/m^3^ to kg/m^3^ for the Asselin site. Then, this decrease was followed by a slight increase until the twelfth year.\

![The ring density versus the number of annual rings of the samples for the Asselin site.The bold lines represent the model prediction](graph_table/Density_asselin.png){width=330px; height=330px} \

![The ring density versus the number of annual rings of the samples for the Saint-Casimir site.The bold lines represent the model prediction](graph_table/Density_casimir.png){width=330px; height=330px} \

@@

|Rank correlation  |  Rho| P-value|
|:-----------------|----:|-------:|
|Variability index | 0.08|    0.43|
|$b_0$             | 0.70|2.2^-16^|
|$b_1$             | 0.80|2.2^-16^|
|$b_2$             | 0.60|2.2^-16^|
Table: Sperman’s Rang correlation

@@

The data set showed a typical radial pattern of wood density already observed in spruce trees in other studies [@Alteyrac2006; @GARDINER2011; @Xiang2014]. The trend is similar to that described as Type II density profile in @panshin1980. The high density during the first years of tree growth can be explained by the immature tracheids combined with the high proportion of compression wood. The density decreases to a certain level and begins to increase again. This rapid decline in density is due to the significant growth of annual rings and the following increase is explained by the fact that the ring width  decreases from a certain age, because of the proportion of earlywood and latewood by ring.[@Lachenbruch2011].\

The modeling of individual profiles is suitable for the description of the wood density of our type II data. The choice of model allowed to take into account the variability of the density profiles of white spruce. @Xiang2014 applied a similar model in their study to describe the density profiles of black spruce with density profiles similar to those of white spruce. \

Although the variaton of density follows a Type II profile, it remains slightly different between the two sites. This difference can be explained by the difference in pedoclimatic conditions between sites. In fact, the soil type is different between sites and the climate is colder in Asselin which is further south compared to Saint-Casimir. At the level of the model parameters, the sites are also different and this difference can be explained by the profiles of densities that are not the same.\

The index of variability is different between two sites, despite the fact that they are the same families of trees. Since the variability index is calculated from the model residuals, the difference can be explained by the fact that the variability that is related to the internal control of the trees has been subtracted from the model and that the remaining one is more controlled by external phenomena related to the study sites. The fact that the index of variability is more under environmental control than genetic explains the lack of heritability of this index. It is not surprising that the variability index is under environmental control because density is mainly controlled by genetics but partly also by growth that is mainly controlled by environmental conditions. The model has made it possible to subtract the typical trend of the variation of the density which is related to the internal functioning of the tree [@Lachenbruch2011] and thus under genetic control. Once the portion of the variation under genetic control is subtracted, the rest of the variability can only be of a pedoclimatic order.\

The index of variability being different between the sites, as one would expect Spearman's rank coefficient of correlation is not significant (p-value = 0.43; rho = 0.08) for this index. This coefficient value is in agreement with the previous results which show a difference between the two sites and the fact that the variability index is under environmental control. The similarity of the correlation coefficient between the three parameters is explained by the fact that the model adapts well to each of the two sites taking into account the extreme variability of the different density profiles.\

Heritability of density is high at both sites. This strong density heritability is similar to previous studies in white spruce [@Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b; @Beaulieu2006]. All of these studies reported values between 0.36 and 0.52. 
In contrast to density, annual ring width has low to moderate heritability in each site and also for combined sites. This result, which is in agreement with previous studies [@Corriveau1991; @Ivkovich2002a], is explained by The difference we observe when we take the sites separately and when they are combined can be explained by a genotype-environment interaction [@kremer1986], this happens when conditions from one environment to another do not have the same effect on the genotype. This result suggests that, if trees are to be selected for reforestation as a result of this study, the territory be divided into relatively homogeneous subunits, within which the recommended sources will be able to keep their superiority and can be used without risk of maladjustment [@beaulieu1989stabilite].

>>>>

Improvement programs by genetic selection are in place for the most important commercial species [@DRF2000; @Mullin2011] including white spruce (Picea glauca (Moench) Voss) which has a significant economic value because of its geographical distribution and abundance [@Eyre1980; @Abrahamson2015]. Spruce wood is valuable as lumber, material for carpentry products and for pulping because of the characteristics of its fibres [@Zhang2008]. In addition, the plasticity of spruce, enable this species to adapt to various types of soil and climatic conditions [@nienstaedt1990]. All these characteristics make white spruce a species of great interest in North America and consequently it is heavily reforested in various silvicultural systems in Canada with more than 150 million of natural an hybrid seedlings planted yearly across almost all forested regions which represents a wide range of soils types and a climatic conditions [@Canadian2009]. \

@@

## Sampling

The material used was sampled from an experimental provenance-progeny test plantation designed to assess the genetic merit of first generation selections, which had been subdivided into breeding groups and delineated by provenance or geographic region. The details of the setting up of the experiment are described in @beaulieu1996. Crosses were made according to a partial diallel mating design within each of the six different breeding groups or sublines [@Burdon1983], which were designed to limit future inbreeding to within group. Each parent tree has been used in crossings one to three times, giving half-brother families (trees with one of two parents in common) and full-brother families (trees with both parents in common) in breeding groups. Genetic testing was conducted in 1999 on 2-year-old nursery seedlings near Asselin balsam fir-yellow birch ecological domain (cool weather), and Saint-Casimir located in the maple–basswood ecological domain (warm weather), Quebec, Canada, in randomized complete blocks to avoid any possible impact of heterogeneity within each site (table 1). Trees were assigned to plots in rows of five trees / plot (spacing of 2 m × 2 m) [@beaulieu2014]. In this study, 93 full-brother families for a total of 2365 trees. The sampled trees were drawn from 18 blocks over the two sites and 9–14 individuals were sampled per full-sib cross. Trees used in this study are exclusively full-brother families.\

|Site          |Location    | LAT (°N)| LON (°W)| ELV (m)| MAT (°C)| TAP (mm)|Soil type        |
|:-------------|:-----------|--------:|--------:|-------:|--------:|--------:|:----------------|
|Asselin       |Squatec     |47.88    |-68.44   |349     |2.4      |1042.0   |loamy soils      |
|Saint-Casimir |Saint-Alban |46.70    |-72.11   |53      |4.7      |1135.7   |sandy-loam soils |
Table: Location and climatic conditions* (over 1981–2010) of the two plantation sites.

*LAT, latitude; LON, longitude; ELV, elevation; MAT, mean annual temperature; TAP, total annual precipitation.

Wood cores were extracted à une épaisseur de 1,68 mm from the south facing side of each tree 15-year-old trees in 2015 with a breast-height diameter (DBH15, 1.3 m above ground). Tree cores are extracted using increment borer which is a specialized tool for extracting wood cores from a living tree with relatively minor damage to the plant itself. These cores were used to determine ring width and wood density. Density was determined by scanning incremental cores from bark to bark along the radial face at a resolution of 25 $\mu$m in the QTRS-01X ring analyzer (Quintek Measurement Systems Inc., Knoxville, TN, USA).  A weighted mean density line was obtained by summing the product between the annual density measurements and the ratio of the respective radial areas to the total radial area. Wood cores were conditioned to about 7% moisture before scanning. Thus, the determination of the density of the wood makes it possible to calculate a quality index of it. From the density profiles, the limits of the annual growth rings were determined together with the relative proportions of earlywood and latewood in each annual ring using in-house data processing routines. Ring width determined through information of densitometry profiles.\

@@

As part of this study, it is essential to accurately identify the year of formation of the annual rings measured. To this end, we used the technique of Cross-dating which consists in obtaining the concordance of the number of rings and of their width pattern to avoid counting errors or those caused by missing rings or by false annual rings. Crossdating is the most basic principle of dendrochronology. Cross-dating is a technique that ensures each individual tree ring is assigned its exact year of formation. This is accomplished by matching patterns of wide and narrow rings between cores from the same tree, and between trees from different locations.
To verify the accuracy of dates and determine if there were missing rings, we used COFECHA, which is a dendrochronology software program used for cross-dating and quality control of annual ring measurements [@HOLMES1983]. Thus, as the annual ring separation was carried out by an algorithm, we eliminated the samples (127 samples) whose dating was not correct and could not be corrected. Any incomplete or false rings were excluded from the dataset.

## Model Development
A model was developed to describe the patterns of variation in mean annual ring density in the stem in white spruce for the purpose of modeling the known density variation portion and subtracting it to better observe the rest of the density variations. Considering the hierarchical structure of the data, with observations on annual rings nested in the cores, cores nested in the trees, and trees nested in blocks, we opted for a mixed-effects modelling approach using nlme package [@NLME2018] in the R statistical programming environment [@R2018] (Eq.3). Previous studies have used this method to describe the density profile of different spruce species that have a similar profile[@franceschini2010; @gardiner2011]. 
However, when modeling density profiles we found that the variation between individual discs follows different patterns from the marrow to the bark between tree. We noticed that it was impossible to optimize the equation using the same estimated parameters for all tree cores, because the pattern of variation of annual rings average density of each tree is more or less different from the others. To work around this problem, we chose to model each of the profiles separately. We, therefore, used the optim function of the stats package in the R [@R2018] statistical programming environment to find the best parameters iteratively for each tree core . The fit of the model was evaluated by visual analysis of the model residual curves against adjusted values and explanatory variables, fit indices (R^2^) [@Pape1999].
As in the study of @Xiang2014, pith-to-bark wood density profiles in our data were different among cores. A nonlinear equation was fit for each core to describe ring density profiles, producing different parameter estimates for each core. The radial pattern of wood density in white spruce can be described as follows:

@@

where RD is the mean ring density (kg.m^-3^), CA is cambial age (years), $b_0$, $b_1$ and $b_2$ are empirically determined parameters, and $e$ is the residual error term. Parameter $b_0$ denotes the intercept when CA approaches a theoretical value of 0.  The graphic description in figure 2 of the three parameters of our model makes it possible to understand how each parameter influences our model (Fig 2).

@@

This modeling of the density profiles then made it possible to subtract the estimated profile which represents the low frequency signal share described previously in the introduction. Thus, only the part of the high-frequency signal that interests us remains to evaluate the variability of the density related to factors external to the tree.\

## Determination of the wood density variability index
After subtraction of the typical density, it is observed that the residues remain variable. Knowing that the low frequency signal is subtracted, this variation can only be due to the high frequency signal which is the interannual variation. As explained earlier in the introduction, low interannual variability is an important criterion for wood quality. In order to identify tree families with less interannual variability in density, we have created a viability index.
The variability index is an index to determine the uniformity in the density of pith-to-bark in order to identify tree families that are least sensitive to variations in climatic conditions. The residues of our model were used to obtain the wood density variability index, which was determined using an equation. We subtracted the density residues values of year $n-1$ from year $n$. We were interested more specifically in the difference between the average density residues of annual rings of two consecutive years, no matter whether the difference was positive or negative. The absolute value was used for each difference to get rid of the sign. The index was estimated from the following formula:

@@

## Spearman’s rank correlation coefficient (Spearman’s rho) 
Spearman’s rank correlation coefficient is a correlation coefficient to describe the link between two variables by means of a monotonic function. We therefore used a Spearman’s rank correlation coefficient to make a comparison of the response of the tree families between the two study sites. We calculated this coefficient for the variability index and the different parameters of the model. The elements for which we calculated Spearman’s rho are: the variability index for each of the tree families between two study sites and the three parameters to estimate from the model. The $cor.test$ function in the R [@R2018] statistical programming environment was used to calculate Spearman’s rho. We used a 95 percent confidence interval.

$$rho = \frac{cov(rg_X,rg_Y)}{\sigma_{rg_X} \times\sigma_{rg_Y}}$$
Where $rho$ is a nonparametric measure of rank correlation, $cov(rg_X,rg_Y)$ is the covariance of the rank variables. $\sigma_{rg_X}$ and $\sigma_{rg_Y}$ are the standard deviations of the rank variables.

@@

![Pith-to-bark radial pattern of breast height density by ring. Fine lines show densities for individual stands of the samples for the Asselin site.The bold lines shows the density predicted by Model](graph_table/Density_asselin.png){width=330px; height=330px}.\

![Pith-to-bark radial pattern of breast height density by ring. Fine lines show densities for individual stands of the samples for the Saint-Casimir site.The bold lines shows the density predicted by Model](graph_table/Density_casimir.png){width=330px; height=330px}. \

When considering all the cores of trees from the pith-to-bark, ring density was high near the pith and decreased rapidly in the first 5 growth rings. It goes from 430 kg/m^3^ to 370 kg/m^3^ for the Saint-Casimir site and from 455kg/m^3^ to kg/m^3^ for the Asselin site. Then, this decrease was followed by a slight increase until the twelfth year. \

@@

|Rank correlation  |  Rho| P-value|
|:-----------------|----:|-------:|
|Density residus   | 0.57|8.7^-09^|
|Variability index | 0.08|    0.43|
|$b_0$             | 0.70|2.2^-16^|
|$b_1$             | 0.80|2.2^-16^|
|$b_2$             | 0.60|2.2^-16^|
Table: Sperman’s Rang correlation

@@

The dataset showed a typical radial pattern of wood density already observed in spruce trees in other researchers [@Alteyrac2006; @GARDINER2011; @Xiang2014] with the density initially decreasing outwards from the pith before slowly increasing towards an asymptotic value. This pattern corresponds to the type II pith-to-bark profile described by @panshin1980. The high density of wood adjacent to the pith has been attributed to high earlywood density and low latewood proportion [@alteyrac2007; @koubaa2007]. In addition, in the first years of a tree's life, it produces "flexure wood" that is high density [@telewski1989]. This type of wood is formed because the low rigidity of the young trees that allows them to withstand environmental constraints without breaking. The proportion of this wood decreases rapidly in the first few rings. The increase in density that follows the decline of the latter is related to the proportion of latewood that increase in mature wood [@clark2007]. Similar hypotheses have been proposed to explain radial variation in wood properties by @Lachenbruch2011. However, this radial pattern does not take into account the interannual variability of density. \

The modeling of individual profiles is suitable for the description of the wood density of our type II data. The prediction of the model on all of the data that can be seen in Figures 3 and 4 is an example of the modeling that is done for individual profiles. The choice of modeling the variation of the density of each of the trees made it possible to adapt the model to the variations between the trees. The fact that the model equation is composed of two components has given sufficient flexibility to properly describe type II density. This type of equation has been used previously in the study of @Xiang2014, where the authors used a three-component equation. In their case, the third component was used to describe a decline near the bark, which is not the case with what is observed in our study.\

Although the density variation to a similar profile, it remains slightly different between the two sites. This difference can be explained by the difference in pedoclimatic conditions between sites. In fact, the soil type is different between sites and the climate is colder in Asselin which is further south compared to Saint-Casimir. In addition, the young age (15 years) of our trees could have an impact on the difference in density profiles between study sites, because it was observed that in white spruce, the environmental control being higher than the genetic control the first years of life of the tree [@Corriveau1991], the conditions of the site thus played an important role. Indeed, the observation of the parameter $b_2$ of model equation allows to see that the rise of the density profile is related to the second term of the equation, and a great value of age cambial causes a greater rise of the profile of density. It can be hypothesised that the effect of the site on density profiles would have been smaller in an older stand. This result is in accordance with observations in white spruce  [@Corriveau1991] and black spruce [@Xiang2014]. At the level of the model parameters, the sites are also different and this difference explained by the profiles of density that are not the same.\

Like the density profiles, variability index is different between the two sites, despite the fact that they are the same families of trees. Since the variability index is calculated from the model residuals, the difference can be explained by the fact that the variability that is related to the internal control of the trees has been subtracted from the model and that the remaining one is more controlled by external phenomena related to the study sites. Variability index being more under environmental control than genetic, this explain the lack of heritability of this index. It is not surprising that the variability index is under environmental control because density is mainly controlled by genetics but partly also by growth that is mainly controlled by environmental conditions. The model has made it possible to subtract the typical pattern of the density variation which is related to the internal functioning of the tree [@Lachenbruch2011] and thus under genetic control. Once the portion of the variation under genetic control is subtracted, the rest of the variability can only be of a pedoclimatic order.\

Variability index being different between the sites, as one would expect Spearman's rank coefficient of correlation is not significant (p-value = 0.43; rho = 0.08) for this index. This coefficient value is in agreement with the previous results which show a difference between the two sites and the fact that the variability index is under environmental control. The similarity of the correlation coefficient between the three parameters is explained by the fact that the model adapts well to each of the two sites taking into account the extreme variability of the different density profiles.\

With regard to heritability of density, it is high at both sites. This strong density heritability is similar to previous studies in white spruce [Nelson 1963 in @mothe1983; @Corriveau1991; @Yanchuk1993; @Ivkovich2002a; @Ivkovich2002b; @Beaulieu2006]. All of these studies reported values between 0.36 and 0.80. 
In contrast to density, annual ring width has low to moderate heritability in each site and also for combined sites. This result, which is in agreement with previous studies [@Corriveau1991; @Ivkovich2002a], is explained by the difference we observe when we take the sites separately and when they are combined can be explained by a genotype-environment interaction [@kremer1986], this happens when conditions from one environment to another do not have the same effect on the genotype. This result suggests that, if trees are to be selected for reforestation as a result of this study, the territory be divided into relatively homogeneous subunits, within which the recommended sources will be able to keep their superiority and can be used without risk of maladjustment [@beaulieu1989stabilite].
